Bound within Boundaries: How Well Do Protected Areas Match Movement Corridors of Their Most Mobile Protected Species?

### 2020-08-21 Chapter_0/04_Manuscript/Manuscript_Backup_2020-06-08.tex
first commit

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
% \usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[nomarkers, nolists]{endfloat}

@@

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Bound within Boundaries: How Well Do Boundaries of Protected Areas Match
Movement Corridors of Their Most Mobile Protected Species?}

% List of Authors
\author{
  David Hofmann\textsuperscript{1\S*} \and
  Dominik Behr\textsuperscript{1,2*} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \and
  Gabriele Cozzi\textsuperscript{1,2}
}

@@

\textsuperscript{*} Shared co-authorship

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
Protecting large portions of natural land connecting natural reserves is an
increasingly used strategy to maintain and restore connectivity among wildlife
subpopulations. One such initiative area is the Kavango-Zambezi Transfrontier
Conservation Area (KAZA-TFCA) in Africa, the world's largest transboundary
conservation area on earth, spanning five countries and 520'000
km\textsuperscript{2}. While boundaries of such conservation areas are often
determined by expert opinion as well as socio-political needs and constraints,
the extent to which these areas match species' movement corridors is rarely
examined. This is mainly due to a lack of adequate empirical data, particularly
on dispersal. Between 2011 and 2019, we collected high-resolution GPS relocation
data on 16 dispersing African wild dog coalitions (\textit{Lycaon pictus}) to
(i) measure differences in landscape permeability within and outside the
boundaries of the KAZA-TFCA and (ii) to assess to which degree the area of the
KAZA-TFCA reflects natural movement corridors of this highly mobile and
endangered species. We show that permeability within the KAZA-TFCA boundaries
was more than double compared to areas outside it. Furthermore, we identify a
five-fold permeability difference among the five KAZA-TFCA countries. These
differences were mainly caused by different degrees of human activities,
hampering dispersal movements. We further show that all major movement corridors
of wild dogs run within the KAZA-TFCA boundaries, although some minor corridors
remain outside formally protected areas. Our work highlights the importance of
empirically assessing the adequacy of protected areas and the need for
international coordination that stem from different contributions to
connectivity across regions and countries.
\end{abstract}

@@

% Create linenumbers
% \linenumbers

\section{Introduction}
As a result of the ongoing human-caused habitat deterioration and fragmentation,
successful management and conservation of isolated subpopulations requires
maintenance and enhancement of connectivity between them \citep{Hanski.1998,
Fahrig.2003, Lindenmayer.2013}. Recently, efforts have been made to facilitate
animal movements and dispersal, and to secure connectivity by protecting
contiguous natural or semi-natural areas \citep{Heller.2009, Doerr.2011,
Rudnick.2012, Cozzi.2020}. These efforts have resulted in the creation of an
ever growing number of large and often transboundary protected areas
\citep{Wolmer.2003}. While boundaries of large scale conservation areas have
been drawn largely based on expert opinion and socio-political needs and
constraints, the usefulness of subjective boundary assessments have revealed
deficiencies in the past \citep{Pullinger.2010}. Due to the very nature of large
scale protected areas, which sometimes span several thousands of square
kilometers and traverse multiple countries, a thorough quantitative assessment
of their effectiveness and adequacy of their boundaries is challenging
\citep{Rudnick.2012}. This challenge is complicated by a lack of animal movement
data at the appropriate spatial scale and an insufficient understanding of the
ranging behavior, movement patterns, and dispersal corridors of the species for
which these protected areas have been created \citep{Vasudev.2015, Tshipa.2017,
Cozzi.2020}. Nevertheless, an accurate representation of movement corridors and
dispersal routes based on empirical global positioning system (GPS) data is
fundamental to objectively assess how well boundaries of protected areas reflect
natural animal movements \citep{Beier.2008, Doerr.2011, Rudnick.2012}.

In recent years, there has been a growing body of research that uses animal
relocation data to identify movement corridors and assess connectivity at large
scales \citep{Chetkiewicz.2006, Doerr.2011, Squires.2013, Elliot.2014,
Benz.2016, Osipova.2019}. Identification of potential corridors relies on the
estimation of permeability (respectively resistance) surfaces, which return the
ease or willingness at which the focal species traverses a specific landscape
\citep{Sawyer.2011}. Such surfaces are typically created on the basis of habitat
preferences, which can be estimated using a suite of selection functions
\citep{Boyce.2002, Fortin.2005, Cushman.2010}. In selection functions, habitat
preferences are inferred by comparing spatial covariates (e.g. environmental and
anthropogenic) at locations visited by the animal to the same spatial covariates
at randomly selected locations in the animal's availability domain
\citep{Zeller.2012, Thurfjell.2014}. Importantly, selection functions rely on
adequate landscape and GPS relocation data that are representative of the
process being studied \citep{Diniz.2020}. More specifically, GPS relocation data
collected on dispersing individuals has been shown to outperform GPS relocation
data collected on resident individuals when aiming at detecting large scale
dispersal corridors \citep{Elliot.2014, Diniz.2020}. However, few connectivity
studies have partitioned their GPS relocation data according to behavioral modes
(e.g. resident vs. dispersing) of the studied species \citep{Wilson.2012,
Vasudev.2015}. As such, most permeability surfaces upon which dispersal
corridors are identified are created using GPS relocation data that were
collected on resident individuals. This introduces severe biases and
substantially reduces the power to reveal meaningful movement corridors, for
dispersing individuals have different needs and drives compared to resident
individuals \citep{Killeen.2014, Elliot.2014, Cozzi.2020}. Consequently, these
biases have limited our ability to assess the adequacy of the boundaries of
protected areas in securing connectivity for their protected species.

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA),
which, spanning over 520'000 km\textsuperscript{2} and five countries,
constitutes the world's largest transfrontier conservation area. While the
KAZA-TFCA was originally set to facilitate movements of African elephants
(\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key to the
conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA's boundaries using
relevant GPS movement data at the appropriate spatial scale \citep{Elliot.2014,
Tshipa.2017}. Thus, how well the KAZA-TFCA's boundaries reflect natural movement
patterns and dispersal corridors of its protected species is virtually unknown.

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly endangered flagship species for conservation efforts, but for which
systematically collected GPS relocation data on dispersing individuals are
unavailable. Once widespread across the entire Sub-Saharan continent, wild
dogs have been widely extirpated through human persecution, habitat destruction,
and disease outbreaks \citep{Woodroffe.2012}. As a result, the species has
become one of Africa's most endangered large carnivores, and currently only
survives in small, spatially scattered subpopulations \citep{Woodroffe.2012}.
Within these subpopulations, wild dogs form cooperative breeding packs of up to
thirty individuals \citep{Frame.1979, Fuller.1992, Creel.2002}, whose social
structure is strongly shaped by the process of dispersal \citep{McNutt.1996,
Woodroffe.2019, Behr.2020}. Both males and females typically disperse from their
natal pack, either alone or in same-sex dispersing coalitions, and search for
unrelated mates and a suitable territory to settle \citep{McNutt.1996,
Cozzi.2020, Behr.2020}. During dispersal, wild dogs can cover several hundred
kilometers, sometimes crossing national boundaries and areas that are strongly
influenced by humans \citep{DaviesMostert.2012, Masenga.2016, Woodroffe.2019,
Cozzi.2020}. Despite the importance of dispersal for the long-term viability of
this species, little empirical information is available on habitat selection and
potential movement barriers of dispersing wild dogs. The few studies that have
collected GPS relocation data during dispersal have shown that dispersers
quickly move over large distances \citep{Woodroffe.2019}, generally avoid
human-dominated landscapes \citep{Masenga.2016, Oneill.2020, Cozzi.2020} and
areas densely forested \citep{Oneill.2020}, but prefer proximity to water and
dirt roads \citep{Oneill.2020}.

Here, we used GPS relocation data collected on 16 dispersing wild dogs in as
many dispersing coalitions from a free-ranging population in northern Botswana
to assess the adequacy of the boundaries of the KAZA-TFCA, the largest protected
area in the world. We estimated the degree of selection or avoidance for
environmental and anthropogenic landscape features. We then applied the obtained
habitat preferences to create a permeability surface spanning the entire
KAZA-TFCA and investigated how landscape permeability varies regionally and
internationally. Finally, we calculated least-cost paths and least-cost
corridors to identify major dispersal corridors, pinpoint dispersal hubs, and
verify whether the KAZA-TFCA successfully covers these features.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
McNutt.2008, Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we
collected GPS relocation data on 16 coalitions of dispersing African wild dogs
(7 female and 9 male coalitions; see also \cite{Abrahms.2017} and
\cite{Cozzi.2020}). Candidate dispersing individuals were identified based on
criteria reported in \cite{Behr.2020}, immobilized according to protocols
described in \cite{Osofsky.1996}, and fitted with GPS/Satellite radio collars
(\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) while still in their
natal pack. All procedures were undertaken and supervised by a
Botswana-registered wildlife veterinarian. Immobilized individuals were
monitored and all were observed to successfully re-join their packs within one
hour after immobilization. During dispersal, GPS collars were programmed to
record GPS relocation data every 4 hours and to regularly transmit them via
Iridium satellite system to a base station. The system thus allowed
remote-tracking of collared individuals, even where field conditions were
prohibitive or dispersal trajectories crossed international borders.

Because we were interested in dispersal behavior, we discarded any GPS data
collected while individuals were still with their natal packs and also after
settlement in a new territory after dispersal \citep{Cozzi.2020}. We identified
the exact time of emigration and settlement based on direct field observations
and through visual inspection of the net squared displacement (NSD) metric. NSD
is a metric returning the euclidean distance of a relocation to a reference
point \citep{Borger.2012}, which in our case was the center of the dispersing
coalition's natal home range. Thus, dispersal was deemed to have started when a
coalition had left its natal home range and continued until the NSD metric
remained stationary, implying that the coalition had settled (Figure S1). Some
coalitions undertook exploratory movements prior to proper dispersal. Such
movements have been observed in other species and are reported to be very
similar to proper dispersal \citep{Killeen.2014}. Therefore, we also classified
exploratory movements as dispersal. In total, we collected 4'169 GPS relocations
during dispersal, i.e. between emigration and settlement (Figure S2 \& Table
S1), resulting in an average of 260 locations per dispersing coalition (range:
37 - 729). On average, coalitions dispersed for 48 days (min = 9, max = 137),
covered a mean euclidean distance of 54 km (min = 3, max = 263) and a mean
cumulative distance of 596 km (min = 130, max = 1'962). In our analysis, we did
not differentiate between male and female dispersing coalitions, for previous
research found little differences between sexes during dispersal
\citep{Woodroffe.2019, Cozzi.2020}.

@@

\subsubsection{Land Cover}
We divided land cover into water and dryland. Water included rivers, wetlands,
and swamps. Because the inundation extent of the flood in the Okavango Delta is
highly variable within and between years, we created dynamic ``flood maps'' that
were updated each 8\textsuperscript{th} day following a remote sensing algorithm
developed by \citep{Wolski.2017}. The algorithm uses thresholding of short
wavelength infrared reflectances of MODIS Terra satellite imagery (MCD43A4;
\citealp{Schaaf.2015}) to distinguish areas covered by water and dryland
(details in Appendix A.3). While we created dynamic flood maps for the Okavango
Delta, we assumed the extent of all other water bodies (e.g. Chobe river,
Zambezi river) to be static within and between years. This static representation
was based on Globeland's land cover dataset \citep{Chen.2015}, from which we
only retained the categories \textit{wetland} and \textit{water bodies} and
collectively reclassified them to \textit{water}. Globeland had an original
resolution of 30m x 30m, so we coarsened the layer to 250m x 250m using the mode
of each 250m x 250m cell. We further improved river representation by employing
the rasterized MERIT Hydro dataset \citep{Yamazaki.2019} from which we added all
rivers with a width of over 10m to our Globeland layer. We merged dynamic and
static water maps into a large rasterstack, covering the entire study area. We
also created a rasterstack rendering the covariate \textit{distance to water} by
calculating the eculidean distance of each raster cell in the study area to the
nearest source of water.

@@

\subsubsection{Protection Status}
We created a binary layer separating protected from unprotected land. We
downloaded corresponding data on protection status from the Peace Parks
Foundation (\url{www.peaceparks.org}; \citealp{PeaceParks.2019}). Protected
areas included forest reserves, game reserves, wildlife management areas, and
national parks. We classified anything not covered by these categories as
unprotected (e.g. communal pastoral land, private land). We rasterized the two
categories to the binary raster \textit{protection status} (1 = protected, 0 =
pastoral) with a resolution of 250m x 250m.

@@

\noindent Because layers (1), (2), and (3) depicted features that are typically
spatially clustered and because not all dispersing coalitions moved within
meaningful distance to each of these features, we totaled the layers
\textit{human density}, \textit{farming}, and \textit{roads}. This resulted in a
single covariate called \textit{human influence}. Note that totalling the layers
implied that the three covariates received different weights (see Appendix A.6
for details). To reduce the influence of outliers, totaled values were limited
to a maximum of 50, which visually resulted in a good balance between high and
low anthropogenic influence and was therefore considered appropriate for our
analysis. To render the fact that humans influence their surroundings beyond
their presence, we followed \cite{Elliot.2014} and applied to each raster cell a
5 km focal buffer within which we summed up and log-transformed human-influence
values (Figure S6).

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' selection or avoidance for the above-mentioned spatial
covariates. In the iSSF framework, GPS relocations are coerced to  ``steps'',
where a step is defined as the connecting line between two consecutive GPS
relocations \citep{Turchin.1998}. Subsequently, habitat preferences are
estimated by contrasting covariates along realized steps to the same covariates
along alternative ``random''  steps \citep{Fortin.2005, Thurfjell.2014,
Avgar.2016}. Thus, we paired each realized step with 24 random steps. We
generated random steps by sampling turning angles from a uniform distribution
U(\(-\pi, +\pi\)) and step lengths from a gamma distribution that was fitted
using realized steps \citep{Avgar.2016}. Together, a realized step and its 24
associated random steps formed a stratum of 25 steps that received a unique
identifier. Along all steps we extracted the above mentioned covariates (see
also Table S3). We also derived a binary covariate indicating whether or not a
step crossed a road and we identified the average distance of each step to the
nearest source of water and the nearest road. We square-rooted the values for
\textit{distance to roads} and \textit{distance to water} to render a decreasing
marginal impact of distance. Additionally, we derived a binary indicator of
whether a step crossed a road or not. We standardized extracted values using a
z-score transformation, and checked for correlation using Pearson's Correlation
Coefficient \(r\). None of the covariates were overly correlated (\(|r| > 0.6\);
\citealp{Latham.2011}) and we retained all of them for modeling. Our habitat
selection was then based on the assumption that dispersing wild dogs assigned a
selection score \(w(x)\) of the following exponential form to each realized and
random step \citep{Fortin.2005, Thurfjell.2014, Avgar.2016}:

>>>>

\documentclass[abstract=on,10pt,a4paper,bibliography=totocnumbered]{article}
\usepackage[paper=a4paper,left=35mm,right=35mm,top=25mm,bottom=30mm]{geometry}
\usepackage[doublespacing]{setspace}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[round]{natbib}
\usepackage{amsmath}
\usepackage{colortbl}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink]{cleveref}
\usepackage{lineno}
\usepackage{multirow}
\usepackage{arydshln}
\usepackage[nomarkers, nolists]{endfloat}

@@

% Change how equations are referenced
\renewcommand{\theequation}{Equation \arabic{equation}}%

%------------------------------------------------------------------------------
%	Titlepage: Header
%------------------------------------------------------------------------------
\title{Bound within Boundaries: How Well Do Protected Areas Match Movement
Corridors of Their Most Mobile Protected Species?}

% List of Authors
\author{
  David D. Hofmann\textsuperscript{1,\S,*} \and
  Dominik M. Behr\textsuperscript{1,2,*} \and
  John W. McNutt\textsuperscript{2} \and
  Arpat Ozgul\textsuperscript{1} \and
  Gabriele Cozzi\textsuperscript{1,2}
}

@@

\textsuperscript{*} Shared first authorship

@@

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
\noindent \textbf{1. Background}: Protecting large portions of natural land
connecting natural reserves is an increasingly used strategy to maintain and
restore connectivity among wildlife subpopulations. One such initiative is the
Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA) in Africa, the
world's largest transboundary conservation area, spanning five countries and
520'000 km\textsuperscript{2}. While boundaries of such conservation areas are
often determined by expert opinion as well as socio-political needs and
constraints, the extent to which they match species' movement corridors is
rarely examined. This is mainly due to a lack of adequate empirical data,
particularly on movement behavior of wide-ranging species during dispersal.

\noindent \textbf{2. Methods}: Between 2011 and 2019, we collected
high-resolution GPS relocation data on 16 coalitions of dispersing African wild
dogs (\textit{Lycaon pictus}) from a free ranging population in northern
Botswana. We combined the collected data with relevant spatial covariates and
used integrated step selection analysis to estimate habitat preferences during
dispersal and create a permeability surface spanning the entire KAZA-TFCA. To
examine whether the KAZA-TFCA covers major movement corridors of this highly
mobile species, we compared landscape permeability across different regions of
our study area. We also calculated least-cost paths and corridors between
selected locations and verified that corridors are encompassed within the
boundaries of the KAZA-TFCA.

\noindent \textbf{3. Results}: We show that permeability within the KAZA-TFCA is
more than double compared to areas outside it. Furthermore, we report a
five-fold permeability difference among the five KAZA-TFCA countries. These
differences are mainly caused by different degrees of human activities across
regions, hampering dispersal movements. We further present a connectivity
network, consisting of least-cost paths and corridors. We thereby show that all
major movement corridors of wild dogs run within the KAZA-TFCA boundaries,
although some minor corridors remain outside formally protected areas.

\noindent \textbf{4. Synthesis and Applications}: Our work highlights the
importance of empirically assessing the adequacy of protected areas and the need
for international coordination that stem from different contributions to
connectivity across regions and countries. Moreover, the presented framework is
readily expandable to include additional species and regions in order to
highlight consistencies among potential dispersal corridors as well as dispersal
corridors outside our ecosystem.

\end{abstract}

@@

% Create linenumbers
\linenumbers

\section{Introduction}
Connectivity among subpopulations is a crucial pre-requisite for many species to
thrive and persist \citep{Fahrig.2003}. A sufficient degree of connectivity
enables dispersal, which in turn contributes to population viability through
gene-flow, the ability to reinforce and rescue small populations, to colonize or
recolonize unoccupied habitats, as well as to shift range in response to climate
change \citep{Brown.1977, Hanski.1998, MacArthur.2001, Frankham.2002,
Heller.2009, Leigh.2012}. Particularly in light of ongoing habitat fragmentation
and destruction worldwide \citep{Fahrig.2003, Lindenmayer.2013}, the protection
of natural or semi natural areas that serve as movement corridors and provide
connectivity has become an utmost task for conservation management
\citep{Heller.2009, Doerr.2011, Rudnick.2012, Cozzi.2020}. This has resulted in
an ever-growing number of large and often transboundary protected areas that are
set to facilitate dispersal between habitats \citep{Wolmer.2003}. Due to the
very nature of large scale protected areas, a thorough quantitative assessment
of their effectiveness and adequacy of their boundaries is challenging
\citep{Rudnick.2012}. In result, boundaries are typically drawn according to
expert opinion and socio-political needs rather than empirical information. In
comparison to qantitative assessments, however, such qualitative approaches have
revealed deficiencies in the past and potentially hold the risk of misdirecting
valuable and scarce conservation funds \citep{Pullinger.2010, Sawyer.2011}. An
accurate representation of movement corridors and dispersal routes based on
empirical data is thus fundamental to objectively assess how well boundaries of
existing or planned protected areas reflect natural movements of their protected
species \citep{Beier.2008, Doerr.2011, Rudnick.2012}.

In recent years, a growing body of research has used animal relocation data to
identify movement corridors and assess connectivity at large scales
\citep{Chetkiewicz.2006, Doerr.2011, Squires.2013, Elliot.2014, Benz.2016,
Osipova.2019}. Identification of potential corridors typically relies on the
estimation of permeability surfaces, which return the ease or willingness at
which the focal species traverses a specific landscape \citep{Sawyer.2011}. Such
surfaces are typically created on the basis of a species' habitat preferences,
which can be quantified using a suite of selection functions \citep{Boyce.2002,
Fortin.2005, Cushman.2010, Zeller.2012}. In selection functions, habitat
preferences are estimated by comparing spatial covariates (e.g. environmental
and anthropogenic) at locations visited by the animal to the same spatial
covariates at randomly selected locations in the animal's availability domain
\citep{Zeller.2012, Thurfjell.2014}. Importantly, selection functions rely on
adequate landscape and relocation data that are representative of the process
being studied \citep{Diniz.2020}. More specifically, relocation data collected
on dispersing individuals has been shown to outperform relocation data collected
on resident individuals, especially when the aim is to detect large scale
dispersal corridors \citep{Elliot.2014, Diniz.2020}. However, dispersal data is
inherently difficult to collect and few connectivity studies have partitioned
their data according to behavioral modes (e.g. resident vs. dispersing) of the
studied species \citep{Wilson.2012, Vasudev.2015}. As such, most permeability
surfaces upon which dispersal corridors are identified are created using
relocation data that were collected on resident individuals. This introduces
severe biases and substantially reduces the power to reveal meaningful movement
corridors, for dispersing individuals have different needs and drives compared
to resident individuals \citep{Killeen.2014, Elliot.2014, Cozzi.2020}.
Consequently, these biases have limited our ability to assess the effectiveness
of protected areas in securing connectivity for their protected species.

One initiative that aims at restoring and enhancing connectivity across large
scales is the Kavango-Zambezi Transfrontier Conservation Area (KAZA-TFCA), which
constitutes the world's largest transfrontier conservation area, spanning over
520'000 km\textsuperscript{2} and five countries (\url{www.kavangozambezi.org}).
While the KAZA-TFCA was originally set to facilitate movements of African
elephants (\textit{Loxodonta africana}; \citealp{Tshipa.2017}), it is also key
to the conservation of other wide-ranging species such as African wild dogs
(\textit{Lycaon pictus}; \citealp{Woodroffe.2012, Cozzi.2020}), lions
(\textit{Panthera leo}; \citealp{Elliot.2014, Cushman.2018}), and cheetahs
(\textit{Acinonyx jubatus}; \citealp{Weise.2017}). To date, however, few studies
have attempted to assess the adequacy of the KAZA-TFCA using relevant global
positioning system (GPS) relocation data of its protected species at the
appropriate spatial scale \citep{Elliot.2014, Tshipa.2017}. Thus, how well the
boundaries of the KAZA-TFCA reflect natural movement patterns and dispersal
corridors of its most mobile protected species is virtually unknown.

Across the KAZA-TFCA, the African wild dog (\textit{Lycaon pictus}) represents a
highly mobile and endangered flagship species for conservation efforts, but for
which systematically collected GPS relocation data on dispersing individuals are
unavailable. Once widespread across the entire Sub-Saharan continent, wild dogs
have been widely extirpated through human persecution, habitat destruction, and
disease outbreaks \citep{Woodroffe.2012}. As a result, the species has become
one of Africa's most endangered large carnivores, and currently only survives in
small, spatially scattered subpopulations \citep{Woodroffe.2012}. Within these
subpopulations, wild dogs form cooperative breeding packs of up to thirty
individuals \citep{Frame.1979, Fuller.1992, Creel.2002}, whose social structure
is strongly governed the process of dispersal \citep{McNutt.1996,
Woodroffe.2019, Behr.2020}. Both males and females disperse from their natal
pack, either alone or in same-sex dispersing coalitions, and search for
unrelated mates and a suitable territory to settle \citep{McNutt.1996,
Cozzi.2020, Behr.2020}. During dispersal, wild dogs can cover several hundred
kilometers, sometimes crossing national boundaries and areas that are strongly
influenced by humans \citep{DaviesMostert.2012, Masenga.2016, Woodroffe.2019,
Cozzi.2020}. Despite the importance of dispersal for the long-term viability of
this species, little empirical information is available on habitat selection and
potential movement barriers during dispersal. The few studies that have
collected GPS relocation data during dispersal have shown that dispersers
quickly move over large distances \citep{Woodroffe.2019}, generally avoid
human-dominated landscapes \citep{Masenga.2016, Oneill.2020, Cozzi.2020} and
areas densely forested \citep{Oneill.2020}, but prefer proximity to water and
dirt roads \citep{Oneill.2020}.

Here, we collected GPS relocation data on 16 dispersing wild dogs in as many
dispersing coalitions from a free-ranging population in northern Botswana and
used it to assess the adequacy of the KAZA-TFCA in securing connectivity for
this species. To achieve this, we estimated the degree of selection or avoidance
for environmental and anthropogenic landscape features during dispersal. We then
applied the obtained habitat preferences and predicted a permeability surface
spanning the entire KAZA-TFCA and we investigated how landscape permeability
varies regionally and internationally. Finally, we calculated least-cost paths
and corridors to identify major movement corridors, pinpoint dispersal hubs, and
verify that the KAZA-TFCA successfully covers these features.

@@

\subsection{GPS Relocation Data}
We used a population of free-ranging African wild dogs inhabiting the Okavango
Delta in northern Botswana as a source population for dispersing individuals.
This population has been extensively studied since 1989 \citep{McNutt.1996,
McNutt.2008, Cozzi.2013, Cozzi.2020, Behr.2020}. Between 2011 and 2019, we
systematically collected GPS relocation data on 16 coalitions of dispersing
African wild dogs (7 female and 9 male coalitions; see also \cite{Abrahms.2017},
\cite{Cozzi.2020}, and \cite{Behr.2020}). Candidate dispersing individuals were
identified based on criteria reported in \cite{Behr.2020}, immobilized according
to protocols described in \cite{Osofsky.1996}, and fitted with GPS/Satellite
radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH, Berlin}) while
still in their natal pack. All procedures were undertaken and supervised by a
Botswana-registered wildlife veterinarian. Immobilized individuals were
monitored and all were observed to successfully re-join their packs within one
hour after immobilization. During dispersal, GPS collars were programmed to
record GPS relocation data every 4 hours and to regularly transmit them via
Iridium satellite system to a base station. The system thus allowed
remote-tracking of collared individuals, even where field conditions were
prohibitive or dispersal trajectories crossed international borders.

Because we were interested in dispersal behavior only, we discarded any GPS data
collected while individuals were still with their natal packs and also after
settlement in a new territory after dispersal \citep{Cozzi.2020}. We identified
the exact time of emigration and settlement based on direct field observations
and through visual inspection of the net squared displacement (NSD) metric. NSD
quantifies the euclidean distance of a relocation to a reference point
\citep{Borger.2012}, which in our case was the center of the dispersing
coalition's natal home range. Thus, dispersal was deemed to have started when a
coalition had left its natal home range and continued until the NSD metric
remained stationary, implying that the coalition had successfully settled
(Figure S1). Some coalitions undertook exploratory movements prior to proper
dispersal. Such movements have been observed in other species and are reported
to be very similar to proper dispersal \citep{Killeen.2014}. Therefore, we also
classified exploratory movements as dispersal. In total, we collected 4'169 GPS
relocations during dispersal, i.e. between emigration and settlement (Figure S2
\& Table S1), resulting in an average of 260 locations per dispersing coalition
(min = 37, max = 729). On average, coalitions dispersed for 48 days (min = 9,
max = 137), covered a mean euclidean distance of 54 km (min = 3, max = 263) and
a mean cumulative distance of 596 km (min = 130, max = 1'962). In our analysis,
we did not differentiate between male and female dispersing coalitions, for
previous research found little differences between sexes during dispersal
\citep{Woodroffe.2019, Cozzi.2020}.

@@

\subsubsection{Land Cover}
We divided land cover into water and dryland. Water included rivers, wetlands,
and swamps. Because the inundation extent of the flood in the Okavango Delta is
highly variable within and between years, we created dynamic ``flood maps'' that
were updated every 8\textsuperscript{th} day following a remote sensing
algorithm developed by \cite{Wolski.2017}. The algorithm uses thresholding of
short wavelength infrared reflectances of MODIS Terra satellite imagery
(MCD43A4; \citealp{Schaaf.2015}) to distinguish areas covered by water and
dryland (details in Appendix A.3). While we created dynamic flood maps for the
Okavango Delta, we assumed the extent of all other water bodies (e.g. Chobe
river, Zambezi river) to be static within and between years. This static
representation was based on Globeland's land cover dataset \citep{Chen.2015},
from which we only retained the categories \textit{wetland} and \textit{water
bodies} and collectively reclassified them to \textit{water}. Globeland had an
original resolution of 30m x 30m, so we coarsened the layer to 250m x 250m using
the mode of each 250m x 250m cell. We further improved river representation by
employing the rasterized MERIT Hydro dataset \citep{Yamazaki.2019} from which we
added all rivers with a width of over 10m to our Globeland layer. We merged
dynamic and static water maps into a large rasterstack, covering the entire
study area. We also created a rasterstack rendering the covariate
\textit{distance to water} by calculating the eculidean distance of each raster
cell in the study area to the nearest source of water.

@@

\subsubsection{Protection Status}
We created a binary layer separating protected from unprotected land. We
downloaded corresponding data on protection status in shapefile format from the
Peace Parks Foundation (\url{www.peaceparks.org}; \citealp{PeaceParks.2019}).
Protected areas included forest reserves, game reserves, wildlife management
areas, and national parks. We classified anything not covered by these
categories as unprotected (e.g. communal pastoral land, private land). We
rasterized the two categories to the binary raster \textit{protection status} (1
= protected, 0 = pastoral) with a resolution of 250m x 250m.

@@

\noindent Because layers (1), (2), and (3) depicted features that are typically
spatially clustered and because not all dispersing coalitions moved within
meaningful distance to each of these features, we totaled the layers
\textit{human density}, \textit{farming}, and \textit{roads}. This resulted in a
single covariate called \textit{human influence}. Note that totaling the layers
implied that the three covariates received different weights (see Appendix A.6
for details). To reduce the influence of outliers, totaled values were limited
to a maximum of 50, which visually resulted in a good balance between high and
low anthropogenic influence and was therefore considered appropriate for our
analysis. To render the fact that humans influence their surroundings beyond
their presence, we followed \cite{Elliot.2014} and applied to each raster cell a
5 km focal buffer within which we summed up and log-transformed human-influence
values (Figure S6).

\subsection{Habitat Selection Model}
We used an integrated step selection function (iSSF; \citealp{Avgar.2016}) to
investigate dispersers' selection or avoidance for the above-mentioned spatial
covariates. In the iSSF framework, GPS relocations are coerced to  ``steps'',
where a step is defined as the connecting line between two consecutive GPS
relocations \citep{Turchin.1998}. Subsequently, habitat preferences are
estimated by contrasting covariates along realized steps to the same covariates
along alternative ``random''  steps \citep{Fortin.2005, Thurfjell.2014,
Avgar.2016}. Thus, we paired each realized step with 24 random steps. We
generated random steps by sampling turning angles from a uniform distribution
U(\(-\pi, +\pi\)) and step lengths from a gamma distribution that was fitted
using realized steps \citep{Avgar.2016}. Together, a realized step and its 24
associated random steps formed a stratum of 25 steps that received a unique
identifier. Along all steps we extracted the above mentioned covariates (see
also Table S3). We also derived a binary covariate indicating whether or not a
step crossed a road and we identified the average distance of each step to the
nearest nearest road and the nearest source of water. We square-rooted the
values for \textit{distance to roads} and \textit{distance to water} to render a
decreasing marginal impact of distance. We standardized extracted values using a
z-score transformation, and checked for correlation using Pearson's Correlation
Coefficient \(r\). None of the covariates were overly correlated (\(|r| > 0.6\);
\citealp{Latham.2011}) and we retained all of them for modeling. Our habitat
selection model was then based on the assumption that dispersing wild dogs
assigned a selection score \(w(x)\) of the following exponential form to each
realized and random step \citep{Fortin.2005, Thurfjell.2014, Avgar.2016}:
